---
title: "Dynamic factor analysis"
author: "James Thorson"
output: rmarkdown::html_vignette
#output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Dynamic factor analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Install locally
#  devtools::install_local( R'(C:\Users\James.Thorson\Desktop\Git\dsem)', force=TRUE )
# Build
#  setwd(R'(C:\Users\James.Thorson\Desktop\Git\dsem)'); devtools::build_rmd("vignettes/dynamic_factor_analysis.Rmd")
```

## Dynamic factor analysis

`dsem` is an R package for fitting dynamic structural equation models (DSEMs) with a simple user-interface and generic specification of simultaneous and lagged effects in a non-recursive structure. Here, we highlight how DSEM can be used to implement dynamic factor analysis (DFA).  We specifically replicate analysis using the Multivariate Autoregressive State-Space (MARSS) package, using data that are provided as an example in the MARSS package.

```{r setup, echo=TRUE, message=FALSE}
library(dsem)
library(MARSS)
library(ggplot2)
data( harborSealWA, package="MARSS")

# Define helper function
grab = \(x,name) x[which(names(x)==name)] 

# Define number of factors
# n_factors >= 3 doesn't seem to converge using DSEM or MARSS without penalties
n_factors = 2 
```


## Using MARSS

We first illustrate a DFA model using three factors, fitted using MARSS:

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
# Load data
dat <- t(scale(harborSealWA[,c("SJI","EBays","SJF","PSnd","HC")]))

# DFA with 3 states; used BFGS because it fits much faster for this model
fit_MARSS <- MARSS( dat, 
                    model = list(m=n_factors),
                    form="dfa",
                    method="BFGS",
                    silent = TRUE )

# Plots states using all data
plot(fit_MARSS, plot.type="xtT")

# Plot expectation for data using all data
plot(fit_MARSS, plot.type="fitted.ytT")
```

## Full-rank covariance using DSEM 

In DSEM syntax, we can first fit a saturated (full-covariance) model using the argument `covs`:

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
# Add factors to data
tsdata = ts( cbind(harborSealWA[,c("SJI","EBays","SJF","PSnd","HC")]), start=1978)

# Scale and center (matches with MARSS does when fitting a DFA)
tsdata = scale( tsdata, center=TRUE, scale=TRUE )

# Define SEM
sem = "
  # Random-walk process for variables 
  SJF -> SJF, 1, NA, 1
  SJI -> SJI, 1, NA, 1
  EBays -> EBays, 1, NA, 1
  PSnd -> PSnd, 1, NA, 1
  HC -> HC, 1, NA, 1
"

# Initial fit
mydsem0 = dsem( tsdata = tsdata,
               covs = c("SJF, SJI, EBays, PSnd, HC"),
               sem = sem,
               family = rep("normal", 5),
               control = dsem_control( quiet = TRUE,
                                       run_model = FALSE ) )   

# fix all measurement errors at diagonal and equal
map = mydsem0$tmb_inputs$map
map$lnsigma_j = factor( rep(1,ncol(tsdata)) )

#
mydsem_full = dsem( tsdata = tsdata,
               covs = c("SJF, SJI, EBays, PSnd, HC"),
               sem = sem,
               family = rep("normal", 5),
               control = dsem_control( quiet = TRUE,
                                       map = map ) )
```

We can then define a custom function to plot states:

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
plot_states = function( out,
                        vars=1:ncol(out$tmb_inputs$data$y_tj) ){
  # 
  xhat_tj = as.list(out$sdrep,report=TRUE,what="Estimate")$z_tj[,vars,drop=FALSE]
  xse_tj = as.list(out$sdrep,report=TRUE,what="Std. Error")$z_tj[,vars,drop=FALSE]

  # 
  longform = expand.grid( Year=time(tsdata), Var=colnames(tsdata)[vars] )
  longform$est = as.vector(xhat_tj)
  longform$se = as.vector(xse_tj)
  longform$upper = longform$est + 1.96*longform$se
  longform$lower = longform$est - 1.96*longform$se
  longform$data = as.vector(tsdata[,vars,drop=FALSE])
  
  # 
  ggplot(data=longform) +  #, aes(x=interaction(var,eq), y=Estimate, color=method)) +
    geom_line( aes(x=Year,y=est) ) +
    geom_point( aes(x=Year,y=data), color="blue", na.rm=TRUE ) +
    geom_ribbon( aes(ymax=as.numeric(upper),ymin=as.numeric(lower), x=Year), color="grey", alpha=0.2 ) + 
    facet_wrap( facets=vars(Var), scales="free", ncol=2 )
}
plot_states( mydsem_full )
```

These estimated states follow the data more closely and have smaller estimated confidence intervals. Presumably this occurs because we are using a full-rank covariance so far.

## Reduced-rank factor model with measurement errors

Next, we can specify three factors factors while eliminating additional process error and estimating measurement errors.  This requires us to switch to `gmrf_parameterization = "projection"`, so that we can fit a rank-deficient Gaussian Markov random field:

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
# Add factors to data
tsdata = harborSealWA[,c("SJI","EBays","SJF","PSnd","HC")]
newcols = array( NA,
                 dim = c(nrow(tsdata),n_factors),
                 dimnames = list(NULL,paste0("F",seq_len(n_factors))) )
tsdata = ts( cbind(tsdata, newcols), start=1978)

# Scale and center (matches with MARSS does when fitting a DFA)
tsdata = scale( tsdata, center=TRUE, scale=TRUE )

#
sem = make_dfa( variables = c("SJI","EBays","SJF","PSnd","HC"),
                n_factors = n_factors )

# Initial fit
mydsem0 = dsem( tsdata = tsdata,
               sem = sem,
               family = c( rep("normal",5), rep("fixed",n_factors) ),
               estimate_delta0 = TRUE,
               control = dsem_control( quiet = TRUE,
                                       run_model = FALSE,
                                       gmrf_parameterization = "projection" ) )

# fix all measurement errors at diagonal and equal
map = mydsem0$tmb_inputs$map
map$lnsigma_j = factor( rep(1,ncol(tsdata)) )

# Fix factors to have initial value, and variables to not
map$delta0_j = factor( c(rep(NA,ncol(harborSealWA)-1), 1:n_factors) )

# Fix variables to have no stationary mean except what's predicted by initial value
map$mu_j = factor( rep(NA,ncol(tsdata)) )

# profile "delta0_j" to match MARSS (which treats initial condition as unpenalized random effect)
mydfa = dsem( tsdata = tsdata,
               sem = sem,
               family = c( rep("normal",5), rep("fixed",n_factors) ),
               estimate_delta0 = TRUE,
               control = dsem_control( quiet = TRUE,
                                       map = map,
                                       use_REML = TRUE,
                                       #profile = "delta0_j",
                                       gmrf_parameterization = "projection" ) )

# Plot estimated factors
plot_states( mydfa, vars=5+seq_len(n_factors) )

# Plot estimated variables
plot_states( mydfa, vars=1:5 )
```

This results in similar (but not identical) factor values using MARSS and DSEM.  In particular, DSEM has higher variance in early years.  This likely arises because the default MARSS implementation of DFA includes a penalty of the initial state $\mathbf{x}_0$ with mean zero and variance of $5\mathbf{I}$.  This term presumably provides additional information about the initial year such that MARSS DFA results are not invariant to reversing the order of the data.

To further explore, we can modify the MARSS DFA to eliminate the prior on initial conditions, based on help from Dr. Eli Holmes.  This involves specifying:

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
# Extract internal settings
modmats <-  summary(fit_MARSS$model, silent=TRUE)

# Redefine defaults
modmats$V0 <- matrix(0, n_factors, n_factors )
modmats$x0 <- "unequal" 

# Refit
fit_MARSS2 = MARSS( dat, 
                    model = modmats,
                    silent = TRUE,
                    control = list( abstol = 0.001,
                                    conv.test.slope.tol = 0.01, 
                                    maxit = 1000 ))
```

We can now compare the three options in terms of estimated loadings:

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
# Compare likelihood for MARSS and DSEM
Table = c( "MARSS" = logLik(fit_MARSS), 
           "DSEM" = logLik(mydfa), 
           "MARSS_no_pen" = logLik(fit_MARSS2) )
knitr::kable( Table, digits=3)       
```
which confirms that the MARSS model without a penalty on initial conditions results in the same likelihood as DSEM.  Finally, we can also compare the three options in terms of estimated loadings:

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
Table = cbind( "MARSS" = as.vector(fit_MARSS$par$Z),
       "DSEM" = grab(mydfa$opt$par,"beta_z"),
       "MARSS_no_pen" = as.vector(fit_MARSS2$par$Z) )
rownames(Table) = names(fit_MARSS$coef)[1:nrow(Table)]
knitr::kable( Table, digits=3)       
```

The estimating loadings are similar using DSEM and the MARSS model without initial penalty, except with label switching (where some factors and loadings can be multiplied by -1 with no change in teh model):

