---
title: "dsem model description"
author: "James T. Thorson"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{dsem model description}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  error = FALSE,
  message = FALSE,
  warning = FALSE
)
# Install locally
#  devtools::install_local( R'(C:\Users\James.Thorson\Desktop\Git\dsem)', force=TRUE )
# Build and PDF
#  setwd(R'(C:\Users\James.Thorson\Desktop\Git\dsem)'); devtools::build_rmd("vignettes/model-description.Rmd"); rmarkdown::render( "vignettes/model-description.Rmd", rmarkdown::pdf_document())
#
# Recommended workflow:
#  * Open in Rstudio and knit using button there
```

\newcommand{\B}{\mathbf}   <!-- math bold nonitalic -->  
\newcommand{\PL}{\mathrm}   <!-- math nonitalic -->   
\newcommand{\s}{\boldsymbol{s}}

# Dynamic structural equation models

dsem involves specifying a dynamic structural equation model (DSEM).  This DSEM be viewed either:

1. *Weak interpretation*: as an expressive interface to parameterize the correlation among variables, using as many or few parameters as might be appropriate; or
2. *Strong interpretation*: as a structural causal model, allowing predictions about the consequence of counterfactual changes to the system.

We introduce DSEM first from the perspective of a software user (i.e., the interface) and then from the perspective of a statistician (i.e., the equations and their interpretation).

### Viewpoint 1:  Software interface

To specify a DSEM, the user uses *arrow-and-lag notation*, based on *arrow notation* derived from package `sem`.  For example, to specify a first-order autoregressive process with variable $x$ this involves:

```{r echo=TRUE, eval=FALSE}
x -> x, 1, ar1
x <-> x, 0, sd
```

This then estimates a single parameter representing first-order autoregression (represented with a one-headed arrow), as well as the Cholesky decomposition of the exogenous covariance of of model variables (specified with two-headed arrows).  See `?make_dsem_ram` Details section for more details about syntax.

If there were four time-intervals ($T=4$) this would then result in the path matrix:

$$
\B P_{\PL{joint}} = 
\begin{pmatrix}
  0 & 0 & 0 & 0 \\
  \rho & 0 & 0 & 0 \\
  0 & \rho & 0 & 0 \\
  0 & 0 & \rho & 0 
\end{pmatrix}
$$
This joint path matrix then represents the partial effect of each variable and time (column) on each other variable and time (row).

DSEM interactions can be as complicated or simple as desired, and can include:

1.  Latent variables and loops (i.e., they are not restricted to directed acyclic graphs);
2.  Values that are fixed a priori, where the `parameter_name` is provided as `NA` and the starting value that follows is the fixed value;
3.  Values that are mirrored among path coefficients, where the same `parameter_name` is provided for multiple rows of the text file.

The user also specifies a distribution for measurement errors for each variable using arguement `family`, and whether each time-series starts from its stationary distribution or from some non-equilibrium initial condition using argument `estimate_delta0`.  If the latter is specified, then variables will tend to converge back on the stationary distribution at a rate that is determined by estimated parameters.  

### Viewpoint 2:  Mathematical details

The DSEM defines a generalized linear mixed model (GLMM) for a $T \times J$ matrix $\mathbf{Y}$, where $y_{tj}$ is the measurement in time $t$ for variable $j$.  This measurement matrix can include missing values $y_{tj} = \PL{NA}$, and it will estimate a $T \times J$ matrix of latent states $\mathbf{X}$ for all modeled times and variables.  

DSEM includes multiple distribution for measurement errors.  For example, if the user specifies `family[j] = "fixed"` then:

$$
y_{tj} = x_{tj} + d_{tj}
$$
for all years.  Alternatively, if the user specifies `family[j] = "normal"` then:
$$
y_{tj} \sim \PL{Normal}( x_{tj} + d_{tj}, {\sigma_j}^2)
$$
and ${\sigma_j}^2$ is then included as an estimated parameter.  These expressions include the $T \times J$ matrix $\B D$ representing the ongoing impact of initial conditions $d_{tj}$ for each variable and year, as explained in detail below.

The specified DSEM then results in Gaussian Markov random field for latent states:

$$
\PL{vec}(\B X) \sim \PL{GMRF}(\B{0, Q}_{\PL{joint}})
$$
where $\B Q_{\PL{joint}}$ is a $TJ \times TJ$ precision matrix such that $\B {Q_{\PL{joint}}}^{-1}$ is the estimated covariance among latent states.  This joint precision is itself constructed from a joint path matrix $\B P_{\PL{joint}}$ and a joint matrix of exogenous covariance $\B G_{\PL{joint}}$:


$$
\B Q_{\PL{joint}} = ({\mathbf{I - P}_{\PL{joint}}}^T) \B G_{\PL{joint}}^{-1} \B G_{\PL{joint}}^{-T} (\mathbf{I - P_{\PL{joint}}})
$$

### Constructing the joint path matrix

The joint path matrix is itself constructed by summing across lagged and simultaneous effects.  Say we specify a model with $K=2$ one-headed arrows.  For each one-headed arrow, we define a $J \times J$ path matrix $\B P_k$ and a lag matrix $\B L_k$.  For example, in a model with $J=3$ variables $(A,B,C)$ and $T=4$ times, and specifying $K=2$ one-headed arrows:

```{r echo=TRUE, eval=FALSE}
A -> B, 0, b_AB
B -> C, 1, b_BC
```

this then results two path matrices:

$$
\B P_1 = 
\begin{pmatrix}
  0 & 0 & 0 \\
  b_{AB} & 0 & 0 \\
  0 & 0 & 0 \\
\end{pmatrix}
$$
and 
$$
\B P_2 = 
\begin{pmatrix}
  0 & 0 & 0 \\
  0 & 0 & 0 \\
  0 & b_{BC} & 0 \\
\end{pmatrix}
$$
with corresponding lag matrices

$$
\B L_{1} = 
\begin{pmatrix}
  1 & 0 & 0 & 0 \\
  0 & 1 & 0 & 0 \\
  0 & 0 & 1 & 0 \\
  0 & 0 & 0 & 1 
\end{pmatrix}
$$
and 
$$
\B L_{2} = 
\begin{pmatrix}
  0 & 0 & 0 & 0 \\
  1 & 0 & 0 & 0 \\
  0 & 1 & 0 & 0 \\
  0 & 0 & 1 & 0 
\end{pmatrix}
$$
We then sum across the Kronecker product of these components to obtain the joint path matrix:

$$
\B P_{\PL{joint}} = \sum_{k=1}^{K}(\B L_k \otimes \B P_k)
$$
where $\otimes$ is the Kronecker product.  Similarly, the exogenous covariance is similar constructed from a Kronecker product, although we assume that all covariance is simultaneous (i.e., no lags are allowed for two-headed arrows):

$$
\B G_{\PL{joint}} = \B I \otimes \B G
$$

These matrices are define a simultaneous equation model:

$$
\PL{vec}(\B X) = \B P_{\PL{joint}} \PL{vec}(\B X) + \B \epsilon \\
\B \epsilon \sim \mathrm{MVN}( \B 0, \B {G_{\PL{joint}}}^T \B G_{\PL{joint}} )
$$
and the precision matrix can be derived from this simultaneous equation. 

Finally, the matrix $\B D$ represents the ongoing effect of initial conditions.  It is constructed from a $J$ length vector of estimated initial conditions $\B \delta_1$ in time $t=1$, and we construct a $T \times J$ matrix $\B \Delta$ where the first row (corresponding to year $t=1$) is $\B \delta_1$ and all other elements are $0$.  The ongoing effect of initial conditions can then be calculated as:

$$
\PL{vec}(\B D) = (\mathbf{I - P}_{\PL{joint}})^{-1} \PL{vec}(\B \Delta)
$$
Calculating the effect $\B D$ of initial conditions requires inverting $\mathbf{I - P}_{\PL{joint}}$, but this is computationally efficient using a sparse LU decomposition.  In general, we can calculate any total effect using this matrix inverse $(\mathbf{I - P}_{\PL{joint}})^{-1}$, and calculating the effect of initial conditions is in a sense the total effect of matrix $\B \delta_1$ in year $t=1$ on subsequent years.  


