---
title: "Spatial diffusion"
author: "James Thorson"
output: rmarkdown::html_vignette
#output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Spatial diffusion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Install locally
#  devtools::install_local( R'(C:\Users\James.Thorson\Desktop\Git\dsem)', force=TRUE )
# Build
#  setwd(R'(C:\Users\James.Thorson\Desktop\Git\dsem)'); devtools::build_rmd("vignettes/dynamic_factor_analysis.Rmd")
```

## Approximating diffusive movement among adjacent spatial strata

`dsem` can be specified to represent diffusion among variables that represent spatial strata where some strata are adjacent to one another.

To show this, we simulate data for five spatial strata (A, B, C, D, E) that are adjacent to one another along a single line (e.g., C is adjacent to B and D): 

```{r simulate, echo=TRUE, message=FALSE}
library(igraph)
library(Matrix)

# Simulation
adjacency_graph = make_graph( ~ A - B - C - D - E )
A = as.matrix( adjacency_graph )

# Diffusion rate
Dprime = 0.2 * A
diag(Dprime) = -1 * rowSums(Dprime)

# Movement transition matrix
M = expm( Dprime )

# 
x_ti = matrix( NA, nrow=100, ncol = nrow(M) )
x_ti[1,] = rnorm(n=nrow(M), mean = 0, sd = 1 )
for( t in 2:nrow(x_ti) ){
  x_ti[t,] = (x_ti[t-1,] %*% M)[1,] + rnorm(n=nrow(M), mean = 0, sd = 0.1)
}
```

We then specify a SEM that approximates diffusive movement, specifically using a diffusion-enhanced spatio-temporal process:

```{r fit, echo=TRUE, message=FALSE}
library(dsem)

# Specify SEM
sem = "
  # Spatial correlation
  A -> B, 0, d0
  B -> C, 0, d0
  C -> D, 0, d0
  D -> E, 0, d0
  E -> D, 0, d0
  D -> C, 0, d0
  C -> B, 0, d0
  B -> A, 0, d0

  # Spatio-temporal diffusion
  A -> B, 1, d
  B -> C, 1, d
  C -> D, 1, d
  D -> E, 1, d
  E -> D, 1, d
  D -> C, 1, d
  C -> B, 1, d
  B -> A, 1, d

  # Self-limitation
  A -> A, 1, rho
  B -> B, 1, rho
  C -> C, 1, rho
  D -> D, 1, rho
  E -> E, 1, rho
"

# Fit
colnames(x_ti) = c("A","B","C","D","E")
fit = dsem(
  tsdata = ts(x_ti),
  sem = sem
)
```

Finally, we can predict movement resulting from the estimated path coefficients:
```{r predict, echo=TRUE, message=FALSE}
# Calculate total effect
effect = total_effect(fit, n_lags = 2)

# Calculate predicted movement
Mhat = array( subset(effect,lag==1)$effect, dim(M) )

# Display predicted movement
knitr::kable( round(Mhat), digits=3)
```

And we can compare this with the true transition matrix from diffusive movement
```{r display, echo=TRUE, message=FALSE}
knitr::kable( round(M), digits=2)
```
