---
title: "Demonstration of selected features"
author: "James Thorson"
output: rmarkdown::html_vignette
#output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Demonstration of selected features}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Install locally
#  devtools::install_local( R'(C:\Users\James.Thorson\Desktop\Git\dsem)', force=TRUE )
# Build
#  setwd(R'(C:\Users\James.Thorson\Desktop\Git\dsem)'); devtools::build_rmd("vignettes/vignette.Rmd"); rmarkdown::render( "vignettes/vignette.Rmd", rmarkdown::pdf_document())
```

```{r setup, echo=TRUE, message=FALSE, warning=FALSE}
library(dsem)
library(dynlm)
library(ggplot2)
library(reshape)
library(gridExtra)
library(phylopath)
```

`dsem` is an R package for fitting dynamic structural equation models (DSEMs) with a simple user-interface and generic specification of simultaneous and lagged effects in a non-recursive structure.   We here highlight a few features in particular.

## Comparison with linear models

We first show that `dsem` is identical to a linear model.  To do so, we simulate data with a single response and single predictor:

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
# simulate normal distribution
x = rnorm(100)
y = 1 + 0.5 * x + rnorm(100)
data = data.frame(x=x, y=y)

# Fit as linear model
Lm = lm( y ~ x, data=data )

# Fit as DSEM
fit = dsem( sem = "x -> y, 0, beta",
            tsdata = ts(data),
            control = dsem_control(quiet=TRUE) )

# Display output
m1 = rbind(
  "lm" = summary(Lm)$coef[2,1:2],
  "dsem" = summary(fit)[1,9:10]
)
knitr::kable( m1, digits=3)
```

This shows that linear and dynamic structural equation models give identical estimates of the single path coefficient.

We can also calculate leave-one-out residuals and display them using DHARMa:

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
# sample-based quantile residuals displayed using DHARMa
samples = loo_residuals(fit, what="samples")
which_use = which(!is.na(data))
fittedPredictedResponse = loo_residuals(fit, what="loo")
res = DHARMa::createDHARMa( simulatedResponse = apply(samples,MARGIN=3,FUN=as.vector)[which_use,],
                            observedResponse = unlist(data)[which_use],
                            fittedPredictedResponse = fittedPredictedResponse$est )
plot(res)
```

This shows no pattern in the quantile-quantile plot, as expected given that we have a correctly specified model

## Comparison with dynamic linear models

We first demonstrate that `dsem` gives identical results to `dynlm` for a well-known econometric model, the Klein-1 model.

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
data(KleinI, package="AER")
TS = ts(data.frame(KleinI, "time"=time(KleinI) - 1931))

# dynlm
fm_cons <- dynlm(consumption ~ cprofits + L(cprofits) + I(pwage + gwage), data = TS)
fm_inv <- dynlm(invest ~ cprofits + L(cprofits) + capital, data = TS)                 #
fm_pwage <- dynlm(pwage ~ gnp + L(gnp) + time, data = TS)

# dsem
sem = "
  # Link, lag, param_name
  cprofits -> consumption, 0, a1
  cprofits -> consumption, 1, a2
  pwage -> consumption, 0, a3
  gwage -> consumption, 0, a3

  cprofits -> invest, 0, b1
  cprofits -> invest, 1, b2
  capital -> invest, 0, b3

  gnp -> pwage, 0, c2
  gnp -> pwage, 1, c3
  time -> pwage, 0, c1
"
tsdata = TS[,c("time","gnp","pwage","cprofits",'consumption',
               "gwage","invest","capital")]
fit = dsem( sem=sem,
            tsdata = tsdata,
            estimate_delta0 = TRUE,
            control = dsem_control(
              quiet = TRUE,
              newton_loops = 0) )

# Compile
m1 = rbind( summary(fm_cons)$coef[-1,],
            summary(fm_inv)$coef[-1,],
            summary(fm_pwage)$coef[-1,] )[,1:2]
m2 = summary(fit$sdrep)[1:9,]
m = rbind(
  data.frame("var"=rownames(m1),m1,"method"="OLS","eq"=rep(c("C","I","Wp"),each=3)),
  data.frame("var"=rownames(m1),m2,"method"="GMRF","eq"=rep(c("C","I","Wp"),each=3))
)
m = cbind(m, "lower"=m$Estimate-m$Std..Error, "upper"=m$Estimate+m$Std..Error )

# ggplot estimates

longform = melt( as.data.frame(KleinI) )
  longform$year = rep( time(KleinI), 9 )
p1 = ggplot( data=longform, aes(x=year, y=value) ) +
  facet_grid( rows=vars(variable), scales="free" ) +
  geom_line( )

p2 = ggplot(data=m, aes(x=interaction(var,eq), y=Estimate, color=method)) +
  geom_point( position=position_dodge(0.9) ) +
  geom_errorbar( aes(ymax=as.numeric(upper),ymin=as.numeric(lower)),
                 width=0.25, position=position_dodge(0.9))  #

p3 = plot( as_fitted_DAG(fit) ) +
     expand_limits(x = c(-0.2,1) )
p4 = plot( as_fitted_DAG(fit, lag=1), text_size=4 ) +
     expand_limits(x = c(-0.2,1), y = c(-0.2,0) )

p1
p2
grid.arrange( arrangeGrob(p3, p4, nrow=2) )
```

Results show that both packages provide (almost) identical estimates and standard errors.

We can also compare results using the Laplace approximation against those obtained via numerical integration of random effects using MCMC.  In this example, MCMC results in somewhat higher estimates of exogenous variance parameters (presumably because those follow a chi-squared distribution with positive skewness), but otherwise the two produce similar estimates.

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=5, eval=FALSE}
library(tmbstan)

# MCMC for both fixed and random effects
mcmc = tmbstan( fit$obj, init="last.par.best" )
summary_mcmc = summary(mcmc)
```
```{r, echo=FALSE, message=FALSE, fig.width=7, fig.height=5, eval=FALSE}
saveRDS( summary_mcmc, file=file.path(R'(C:\Users\James.Thorson\Desktop\Git\dsem\inst\tmbstan)',"summary_mcmc.RDS") )
```
```{r, echo=FALSE, message=FALSE, fig.width=6, fig.height=4, out.width = "100%", eval=TRUE}
summary_mcmc = readRDS( file.path(system.file("tmbstan",package="dsem"),"summary_mcmc.RDS") )
```
```{r, echo=TRUE, message=FALSE, fig.width=6, fig.height=4, out.width = "100%", eval=TRUE}
# long-form data frame
m1 = summary_mcmc$summary[1:17,c('mean','sd')]
rownames(m1) = paste0( "b", seq_len(nrow(m1)) )
m2 = summary(fit$sdrep)[1:17,c('Estimate','Std. Error')]
m = rbind(
  data.frame('mean'=m1[,1], 'sd'=m1[,2], 'par'=rownames(m1), "method"="MCMC"),
  data.frame('mean'=m2[,1], 'sd'=m2[,2], 'par'=rownames(m1), "method"="LA")
)
m$lower = m$mean - m$sd
m$upper = m$mean + m$sd

# plot
ggplot(data=m, aes(x=par, y=mean, col=method)) +
  geom_point( position=position_dodge(0.9) ) +
  geom_errorbar( aes(ymax=as.numeric(upper),ymin=as.numeric(lower)),
                 width=0.25, position=position_dodge(0.9))  #
```

## Comparison with vector autoregressive models

We next demonstrate that `dsem` gives similar results to a vector autoregressive (VAR) model.  To do so, we analyze population abundance of wolf and moose populations on Isle Royale from 1959 to 2019, downloaded from their website (Vucetich, JA and Peterson RO. 2012. The population biology of Isle Royale wolves and moose: an overview. URL: www.isleroyalewolf.org).

This dataset was previously analyzed by in Chapter 14 of the User Manual for the R-package MARSS (Holmes, E. E., M. D. Scheuerell, and E. J. Ward (2023) Analysis of multivariate time-series using the MARSS package. Version 3.11.8. NOAA Fisheries,
  Northwest Fisheries Science Center, 2725 Montlake Blvd E., Seattle, WA 98112, DOI: 10.5281/zenodo.5781847).

Here, we compare fits using `dsem` with `dynlm`, as well as a vector autoregressive model package `vars`, and finally with `MARSS`.

```{r, echo=TRUE, message=FALSE, fig.width=5, fig.height=7, warning=FALSE}
data(isle_royale)
data = ts( log(isle_royale[,2:3]), start=1959)

sem = "
  # Link, lag, param_name
  wolves -> wolves, 1, arW
  moose -> wolves, 1, MtoW
  wolves -> moose, 1, WtoM
  moose -> moose, 1, arM
"
# initial first without delta0 (to improve starting values)
fit0 = dsem( sem = sem,
             tsdata = data,
             estimate_delta0 = FALSE,
             control = dsem_control(
               quiet = FALSE,
               getsd = FALSE) )

#
parameters = fit0$obj$env$parList()
  parameters$delta0_j = rep( 0, ncol(data) )

# Refit with delta0
fit = dsem( sem = sem,
            tsdata = data,
            estimate_delta0 = TRUE,
            control = dsem_control( quiet=TRUE,
                                    parameters = parameters ) )

# dynlm
fm_wolf = dynlm( wolves ~ 1 + L(wolves) + L(moose), data=data )   #
fm_moose = dynlm( moose ~ 1 + L(wolves) + L(moose), data=data )   #

# MARSS
library(MARSS)
z.royale.dat <- t(scale(data.frame(data),center=TRUE,scale=FALSE))
royale.model.1 <- list(
  Z = "identity",
  B = "unconstrained",
  Q = "diagonal and unequal",
  R = "zero",
  U = "zero"
)
kem.1 <- MARSS(z.royale.dat, model = royale.model.1)
SE <- MARSSparamCIs( kem.1 )

# Using VAR package
library(vars)
var = VAR( data, type="const" )

### Compile
m1 = rbind( summary(fm_wolf)$coef[-1,], summary(fm_moose)$coef[-1,] )[,1:2]
m2 = summary(fit$sdrep)[1:4,]
#m2 = cbind( "Estimate"=fit$opt$par, "Std. Error"=fit$sdrep$par.fixed )[1:4,]
m3 = cbind( SE$parMean[c(1,3,2,4)], SE$par.se$B[c(1,3,2,4)] )
colnames(m3) = colnames(m2)
m4 = rbind( summary(var$varresult$wolves)$coef[-3,], summary(var$varresult$moose)$coef[-3,] )[,1:2]

# Bundle
m = rbind(
  data.frame("var"=rownames(m1), m1, "method"="dynlm", "eq"=rep(c("Wolf","Moose"),each=2)),
  data.frame("var"=rownames(m1), m2, "method"="dsem", "eq"=rep(c("Wolf","Moose"),each=2)),
  data.frame("var"=rownames(m1), m3, "method"="MARSS", "eq"=rep(c("Wolf","Moose"),each=2)),
  data.frame("var"=rownames(m1), m4, "method"="vars", "eq"=rep(c("Wolf","Moose"),each=2))
)
#knitr::kable( m1, digits=3)
#knitr::kable( m2, digits=3)

m = cbind(m, "lower"=m$Estimate-m$Std..Error, "upper"=m$Estimate+m$Std..Error )

# ggplot estimates ... interaction(x,y) causes an error sometimes
library(ggplot2)
library(ggpubr)
library(ggraph)
longform = reshape( isle_royale, idvar = "year", direction="long", varying=list(2:3), v.names="abundance", timevar="species", times=c("wolves","moose") )
p1 = ggplot( data=longform, aes(x=year, y=abundance) ) +
  facet_grid( rows=vars(species), scales="free" ) +
  geom_point( )

p2 = ggplot(data=m, aes(x=interaction(var,eq), y=Estimate, color=method)) +
  geom_point( position=position_dodge(0.9) ) +
  geom_errorbar( aes(ymax=as.numeric(upper),ymin=as.numeric(lower)),
                 width=0.25, position=position_dodge(0.9))  #
p3 = plot( as_fitted_DAG(fit, lag=1), rotation=0 ) +
     geom_edge_loop( aes( label=round(weight,2), direction=0)) + #arrow=arrow(), , angle_calc="along", label_dodge=grid::unit(10,"points") )
     expand_limits(x = c(-0.1,0) )

ggarrange( p1, p2, p3,
           labels = c("Time-series data", "Estimated effects", "Fitted path digram"),
           ncol = 1, nrow = 3)
```

Results again show that `dsem` can estimate parameters for a vector autoregressive model (VAM), and it exactly matches results from `vars`, using `dynlm`, or using `MARSS`.

## Multi-causal ecosystem synthesis

We next replicate an analysis involving climate, forage fishes, stomach contents, and recruitment of a predatory fish.

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
data(bering_sea)
Z = ts( bering_sea )
family = rep('fixed', ncol(bering_sea))

# Specify model
sem = "
  # Link, lag, param_name
  log_seaice -> log_CP, 0, seaice_to_CP
  log_CP -> log_Cfall, 0, CP_to_Cfall
  log_CP -> log_Esummer, 0, CP_to_E
  log_PercentEuph -> log_RperS, 0, Seuph_to_RperS
  log_PercentCop -> log_RperS, 0, Scop_to_RperS
  log_Esummer -> log_PercentEuph, 0, Esummer_to_Suph
  log_Cfall -> log_PercentCop, 0, Cfall_to_Scop
  SSB -> log_RperS, 0, SSB_to_RperS

  log_seaice -> log_seaice, 1, AR1, 0.001
  log_CP -> log_CP, 1,  AR2, 0.001
  log_Cfall -> log_Cfall, 1, AR4, 0.001
  log_Esummer -> log_Esummer, 1, AR5, 0.001
  SSB -> SSB, 1, AR6, 0.001
  log_RperS ->  log_RperS, 1, AR7, 0.001
  log_PercentEuph -> log_PercentEuph, 1, AR8, 0.001
  log_PercentCop -> log_PercentCop, 1, AR9, 0.001
"

# Fit
fit = dsem( sem = sem,
            tsdata = Z,
            family = family,
            control = dsem_control(use_REML=FALSE, quiet=TRUE) )
ParHat = fit$obj$env$parList()
# summary( fit )
```
```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
# Timeseries plot
oldpar <- par(no.readonly = TRUE)
par( mfcol=c(3,3), mar=c(2,2,2,0), mgp=c(2,0.5,0), tck=-0.02 )
for(i in 1:ncol(bering_sea)){
  tmp = bering_sea[,i,drop=FALSE]
  tmp = cbind( tmp, "pred"=ParHat$x_tj[,i] )
  SD = as.list(fit$sdrep,what="Std.")$x_tj[,i]
  tmp = cbind( tmp, "lower"=tmp[,2] - ifelse(is.na(SD),0,SD),
                    "upper"=tmp[,2] + ifelse(is.na(SD),0,SD) )
  #
  plot( x=rownames(bering_sea), y=tmp[,1], ylim=range(tmp,na.rm=TRUE),
        type="p", main=colnames(bering_sea)[i], pch=20, cex=2 )
  lines( x=rownames(bering_sea), y=tmp[,2], type="l", lwd=2,
         col="blue", lty="solid" )
  polygon( x=c(rownames(bering_sea),rev(rownames(bering_sea))),
           y=c(tmp[,3],rev(tmp[,4])), col=rgb(0,0,1,0.2), border=NA )
}
par(oldpar)
```
```{r, echo=TRUE, message=FALSE, fig.width=8, fig.height=8}
#
library(phylopath)
library(ggplot2)
library(ggpubr)
library(reshape)
library(gridExtra)
longform = melt( bering_sea )
  longform$year = rep( 1963:2023, ncol(bering_sea) )
p0 = ggplot( data=longform, aes(x=year, y=value) ) +
  facet_grid( rows=vars(variable), scales="free" ) +
  geom_point( )

p1 = plot( (as_fitted_DAG(fit)), edge.width=1, type="width",
           text_size=4, show.legend=FALSE,
           arrow = grid::arrow(type='closed', 18, grid::unit(10,'points')) ) +
     scale_x_continuous(expand = c(0.4, 0.1))
p1$layers[[1]]$mapping$edge_width = 1
p2 = plot( (as_fitted_DAG(fit, what="p_value")), edge.width=1, type="width",
           text_size=4, show.legend=FALSE, colors=c('black', 'black'),
           arrow = grid::arrow(type='closed', 18, grid::unit(10,'points')) ) +
     scale_x_continuous(expand = c(0.4, 0.1))
p2$layers[[1]]$mapping$edge_width = 0.5

#grid.arrange( arrangeGrob( p0+ggtitle("timeseries"),
#              arrangeGrob( p1+ggtitle("Estimated path diagram"),
#                           p2+ggtitle("Estimated p-values"), nrow=2), ncol=2 ) )
ggarrange(p1, p2, labels = c("Simultaneous effects", "Two-sided p-value"),
                    ncol = 1, nrow = 2)
```

These results are further discussed in the paper describing dsem.

## Site-replicated trophic cascade

Finally, we replicate an analysis involving a trophic cascade involving sea stars predators, sea urchin consumers, and kelp producers.

```{r, echo=TRUE, message=FALSE, fig.width=5, fig.height=7, warning=FALSE}
data(sea_otter)
Z = ts( sea_otter[,-1] )

# Specify model
sem = "
  Pycno_CANNERY_DC -> log_Urchins_CANNERY_DC, 0, x2
  log_Urchins_CANNERY_DC -> log_Kelp_CANNERY_DC, 0, x3
  log_Otter_Count_CANNERY_DC -> log_Kelp_CANNERY_DC, 0, x4

  Pycno_CANNERY_UC -> log_Urchins_CANNERY_UC, 0, x2
  log_Urchins_CANNERY_UC -> log_Kelp_CANNERY_UC, 0, x3
  log_Otter_Count_CANNERY_UC -> log_Kelp_CANNERY_UC, 0, x4

  Pycno_HOPKINS_DC -> log_Urchins_HOPKINS_DC, 0, x2
  log_Urchins_HOPKINS_DC -> log_Kelp_HOPKINS_DC, 0, x3
  log_Otter_Count_HOPKINS_DC -> log_Kelp_HOPKINS_DC, 0, x4

  Pycno_HOPKINS_UC -> log_Urchins_HOPKINS_UC, 0, x2
  log_Urchins_HOPKINS_UC -> log_Kelp_HOPKINS_UC, 0, x3
  log_Otter_Count_HOPKINS_UC -> log_Kelp_HOPKINS_UC, 0, x4

  Pycno_LOVERS_DC -> log_Urchins_LOVERS_DC, 0, x2
  log_Urchins_LOVERS_DC -> log_Kelp_LOVERS_DC, 0, x3
  log_Otter_Count_LOVERS_DC -> log_Kelp_LOVERS_DC, 0, x4

  Pycno_LOVERS_UC -> log_Urchins_LOVERS_UC, 0, x2
  log_Urchins_LOVERS_UC -> log_Kelp_LOVERS_UC, 0, x3
  log_Otter_Count_LOVERS_UC -> log_Kelp_LOVERS_UC, 0, x4

  Pycno_MACABEE_DC -> log_Urchins_MACABEE_DC, 0, x2
  log_Urchins_MACABEE_DC -> log_Kelp_MACABEE_DC, 0, x3
  log_Otter_Count_MACABEE_DC -> log_Kelp_MACABEE_DC, 0, x4

  Pycno_MACABEE_UC -> log_Urchins_MACABEE_UC, 0, x2
  log_Urchins_MACABEE_UC -> log_Kelp_MACABEE_UC, 0, x3
  log_Otter_Count_MACABEE_UC -> log_Kelp_MACABEE_UC, 0, x4

  Pycno_OTTER_PT_DC -> log_Urchins_OTTER_PT_DC, 0, x2
  log_Urchins_OTTER_PT_DC -> log_Kelp_OTTER_PT_DC, 0, x3
  log_Otter_Count_OTTER_PT_DC -> log_Kelp_OTTER_PT_DC, 0, x4

  Pycno_OTTER_PT_UC -> log_Urchins_OTTER_PT_UC, 0, x2
  log_Urchins_OTTER_PT_UC -> log_Kelp_OTTER_PT_UC, 0, x3
  log_Otter_Count_OTTER_PT_UC -> log_Kelp_OTTER_PT_UC, 0, x4

  Pycno_PINOS_CEN -> log_Urchins_PINOS_CEN, 0, x2
  log_Urchins_PINOS_CEN -> log_Kelp_PINOS_CEN, 0, x3
  log_Otter_Count_PINOS_CEN -> log_Kelp_PINOS_CEN, 0, x4

  Pycno_SIREN_CEN -> log_Urchins_SIREN_CEN, 0, x2
  log_Urchins_SIREN_CEN -> log_Kelp_SIREN_CEN, 0, x3
  log_Otter_Count_SIREN_CEN -> log_Kelp_SIREN_CEN, 0, x4

  # AR1
  Pycno_CANNERY_DC -> Pycno_CANNERY_DC, 1, ar1
  log_Urchins_CANNERY_DC -> log_Urchins_CANNERY_DC, 1, ar2
  log_Otter_Count_CANNERY_DC -> log_Otter_Count_CANNERY_DC, 1, ar3
  log_Kelp_CANNERY_DC -> log_Kelp_CANNERY_DC, 1, ar4

  Pycno_CANNERY_UC -> Pycno_CANNERY_UC, 1, ar1
  log_Urchins_CANNERY_UC -> log_Urchins_CANNERY_UC, 1, ar2
  log_Otter_Count_CANNERY_UC -> log_Otter_Count_CANNERY_UC, 1, ar3
  log_Kelp_CANNERY_UC -> log_Kelp_CANNERY_UC, 1, ar4

  Pycno_HOPKINS_DC -> Pycno_HOPKINS_DC, 1, ar1
  log_Urchins_HOPKINS_DC -> log_Urchins_HOPKINS_DC, 1, ar2
  log_Otter_Count_HOPKINS_DC -> log_Otter_Count_HOPKINS_DC, 1, ar3
  log_Kelp_HOPKINS_DC -> log_Kelp_HOPKINS_DC, 1, ar4

  Pycno_HOPKINS_UC -> Pycno_HOPKINS_UC, 1, ar1
  log_Urchins_HOPKINS_UC -> log_Urchins_HOPKINS_UC, 1, ar2
  log_Otter_Count_HOPKINS_UC -> log_Otter_Count_HOPKINS_UC, 1, ar3
  log_Kelp_HOPKINS_UC -> log_Kelp_HOPKINS_UC, 1, ar4

  Pycno_LOVERS_DC -> Pycno_LOVERS_DC, 1, ar1
  log_Urchins_LOVERS_DC -> log_Urchins_LOVERS_DC, 1, ar2
  log_Otter_Count_LOVERS_DC -> log_Otter_Count_LOVERS_DC, 1, ar3
  log_Kelp_LOVERS_DC -> log_Kelp_LOVERS_DC, 1, ar4

  Pycno_LOVERS_UC -> Pycno_LOVERS_UC, 1, ar1
  log_Urchins_LOVERS_UC -> log_Urchins_LOVERS_UC, 1, ar2
  log_Otter_Count_LOVERS_UC -> log_Otter_Count_LOVERS_UC, 1, ar3
  log_Kelp_LOVERS_UC -> log_Kelp_LOVERS_UC, 1, ar4

  Pycno_MACABEE_DC -> Pycno_MACABEE_DC, 1, ar1
  log_Urchins_MACABEE_DC -> log_Urchins_MACABEE_DC, 1, ar2
  log_Otter_Count_MACABEE_DC -> log_Otter_Count_MACABEE_DC, 1, ar3
  log_Kelp_MACABEE_DC -> log_Kelp_MACABEE_DC, 1, ar4

  Pycno_MACABEE_UC -> Pycno_MACABEE_UC, 1, ar1
  log_Urchins_MACABEE_UC -> log_Urchins_MACABEE_UC, 1, ar2
  log_Otter_Count_MACABEE_UC -> log_Otter_Count_MACABEE_UC, 1, ar3
  log_Kelp_MACABEE_UC -> log_Kelp_MACABEE_UC, 1, ar4

  Pycno_OTTER_PT_DC -> Pycno_OTTER_PT_DC, 1, ar1
  log_Urchins_OTTER_PT_DC -> log_Urchins_OTTER_PT_DC, 1, ar2
  log_Otter_Count_OTTER_PT_DC -> log_Otter_Count_OTTER_PT_DC, 1, ar3
  log_Kelp_OTTER_PT_DC -> log_Kelp_OTTER_PT_DC, 1, ar4

  Pycno_OTTER_PT_UC -> Pycno_OTTER_PT_UC, 1, ar1
  log_Urchins_OTTER_PT_UC -> log_Urchins_OTTER_PT_UC, 1, ar2
  log_Otter_Count_OTTER_PT_UC -> log_Otter_Count_OTTER_PT_UC, 1, ar3
  log_Kelp_OTTER_PT_UC -> log_Kelp_OTTER_PT_UC, 1, ar4

  Pycno_PINOS_CEN -> Pycno_PINOS_CEN, 1, ar1
  log_Urchins_PINOS_CEN -> log_Urchins_PINOS_CEN, 1, ar2
  log_Otter_Count_PINOS_CEN -> log_Otter_Count_PINOS_CEN, 1, ar3
  log_Kelp_PINOS_CEN -> log_Kelp_PINOS_CEN, 1, ar4

  Pycno_SIREN_CEN -> Pycno_SIREN_CEN, 1, ar1
  log_Urchins_SIREN_CEN -> log_Urchins_SIREN_CEN, 1, ar2
  log_Otter_Count_SIREN_CEN -> log_Otter_Count_SIREN_CEN, 1, ar3
  log_Kelp_SIREN_CEN -> log_Kelp_SIREN_CEN, 1, ar4
"

# Fit model
fit = dsem( sem = sem,
            tsdata = Z,
            control = dsem_control(use_REML=FALSE, quiet=TRUE) )
# summary( fit )

#
library(phylopath)
library(ggplot2)
library(ggpubr)
get_part = function(x){
  vars = c("log_Kelp","log_Otter","log_Urchin","Pycno")
  index = sapply( vars, FUN=\(y) grep(y,rownames(x$coef))[1] )
  x$coef = x$coef[index,index]
  dimnames(x$coef) = list( vars, vars )
  return(x)
}
p1 = plot( get_part(as_fitted_DAG(fit)), type="width", show.legend=FALSE)
p1$layers[[1]]$mapping$edge_width = 0.5
p2 = plot( get_part(as_fitted_DAG(fit, what="p_value" )), type="width",
           show.legend=FALSE, colors=c('black', 'black'))
p2$layers[[1]]$mapping$edge_width = 0.1

longform = melt( sea_otter[,-1], as.is=TRUE )
longform$X1 = 1999:2019[longform$X1]
longform$Site = gsub( "log_Kelp_", "",
                gsub( "log_Urchins_", "",
                gsub( "Pycno_", "",
                gsub( "log_Otter_Count_", "", longform$X2))))
longform$Species = sapply( seq_len(nrow(longform)), FUN=\(i)gsub(longform$Site[i],"",longform$X2[i]) )
p3 = ggplot( data=longform, aes(x=X1, y=value, col=Species) ) +
  facet_grid( rows=vars(Site), scales="free" ) +
  geom_line( )

ggarrange(p1 + scale_x_continuous(expand = c(0.3, 0)),
                    p2 + scale_x_continuous(expand = c(0.3, 0)),
                    labels = c("Simultaneous effects", "Two-sided p-value"),
                    ncol = 1, nrow = 2)
```

Again, these results are further discussed in the paper describing dsem.


